---
title: "Open And Render Check"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This notebook checks whether the basic course workflow is working.

By the end, you should know that:

- the project opens correctly
- the setup chunk runs
- R code chunks run
- figures display
- Quarto can render the notebook to HTML

# Check The Project

Confirm that R can see where this notebook is running.

```{r}
getwd()
```

You should be working somewhere inside the course project.

# Check R Code

Run a simple calculation.

```{r}
2 + 2
```

Create a small object.

```{r}
favorite_number <- 42
favorite_number
```

# Check Data

For the first day, we will use the built-in `mtcars` dataset so nothing has to be downloaded.

```{r}
mtcars |>
  head()
```

```{r}
mtcars |>
  summarise(
    rows = n(),
    average_mpg = mean(mpg),
    minimum_mpg = min(mpg),
    maximum_mpg = max(mpg)
  )
```

# Check Plotting

Make a quick plot.

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 5) +
  labs(
    title = "Distribution of Fuel Efficiency",
    x = "Miles per gallon",
    y = "Number of cars"
  )
```

# Render Check

Now render the notebook.

In RStudio or Positron, click the Render button.

After rendering, confirm that:

- an HTML file was created
- the HTML file opens in a browser or viewer
- the plot appears in the rendered file
- the text and code output appear in the rendered file

# Troubleshooting Notes

Common problems:

- The notebook was opened outside the course project.
- R or Quarto is not installed correctly.
- The first code chunk did not run.
- The render failed because an earlier chunk had an error.

When something fails, read the first error message carefully and fix that first.
