---
title: "Practice: Data Cleaning"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice a basic data-cleaning workflow with one data file.

You will practice:

- `select()`
- `rename()`
- `mutate()`
- `filter()`
- `group_by()` and `summarize()`
- reordering categories by a value
- releveling categories in a chosen order

# Data

Use NHANES. I've set it up below.

This file has one row per NHANES participant in the teaching dataset.

# Work

## Load Data

```{r}
# Load the NHANES analysis data.

dat <- readr::read_csv('data/real/nhanes_2021_2023_analysis.csv')

```

## Select Columns

Select this smaller set of variables:

- `respondent_id`
- `gender`
- `age_years`
- `race_ethnicity`
- `pulse_1`
- `pulse_2`
- `pulse_3`
- `adult_bp_category_simplified`

```{r}
# Use select() to keep the columns you want to work with.

dat <- dat |>
  select(
    # ...
  )
```

## Rename Columns

Rename respondent_id to id.

```{r}
# Use rename() to give at least two columns clearer names.

dat <- dat |>
  rename(
    # ... # The format is new_name = old_name
  )
```

## Mutate

Use `mutate()` to create a variable `pulse_mean` and a categorical version of `age_years`.

I started the `case_when()` statement for the age category. Finish the remaining age groups.

```{r}
# Use mutate() to create pulse_mean and age_group.
dat |> select(age_years) |> summary()

dat <- dat |>
  mutate(
    pulse_mean = # ...,
    age_group = case_when(
      age_years < 18 ~ "Under 18",
      age_years >= 18 & age_years <= 44 ~ "18-44",
      # Add the next age group here.
      # Add the oldest age group here.
      TRUE ~ NA_character_ # This acts as a fallback.
    )
  )
```

## Filter

Create a different dataframe filtered to a subset of 18-44 year-olds.

```{r}
# Use filter() to keep the rows you need.

young_adults <- dat |>
  filter(...) |> # fill in here
  filter(...) # Hint: it's easiest with two lines of filtering
```

## Group And Summarize

Create one grouped summary table of average pulse by age group.

```{r}
# Use group_by() and summarize() to create a grouped summary table.

pulse_by_age <- dat |>
  group_by(...) |> # Fill out
  summarize(
    mean_pulse = mean(...),  # fill out... you can  add ", NA.RM = TRUE" to remove NAs
    n = ..., # fill out
    .groups = "drop" # this "ungroups" the data
  )

pulse_by_age
```

## Reorder

Reorder the race_ethnicity category by average pulse.

```{r}
# Use reorder() to order categories by a summary value.

pulse_by_race_ethnicity <- dat |>
  group_by(race_ethnicity) |>
  summarize(
    mean_pulse = mean(pulse_mean, na.rm = TRUE),  # This may look awkward.  But we take a mean of a mean, because we're in a group_by
    n = n(),
    .groups = "drop"
  ) |>
  mutate(
    race_ethnicity = reorder(...) # fill it in. the format is:  variable name, order variable
  )

pulse_by_race_ethnicity
```

## Relevel

Relevel `adult_bp_category_simplified` in this order

1. `normal_range`
2. `elevated_range`
3. `stage_1_range`
4. `stage_2_range`

```{r}
# Use factor(..., levels = c(...)) to set a meaningful category order.

dat <- dat |>
  mutate(
    adult_bp_category_simplified = factor(
      adult_bp_category_simplified,
      levels = c(
        # write the levels out here in quotes, like "normal_range", ...
      )
    )
  )

dat |>
  count(adult_bp_category_simplified)
```
