---
title: "Practice: Continuous Variables"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice a simple continuous-variable workflow:

- load one data file
- choose one numeric variable
- make a short summary
- plot the distribution
- change the bin width and compare
- use a boxplot to compare spread across groups

# Data

Use:

`data/real/nhanes_2021_2023_analysis.csv`

Good variables to try:

- `age_years`
- `systolic_mean`
- `diastolic_mean`
- `pulse_mean`
- `mean_daily_energy_kcal`

Helpful module reference: `modules/04_continuous-data/02_boxplots-violin-plots-and-outliers.qmd`.

# Work

## Load Data

```{r}
# Load the NHANES analysis data.

dat <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")
```

## Select One Numeric Variable

```{r}
# Select one numeric variable and filter out missing values.

one_variable <- dat |>
  select(...) |>
  filter(!is.na(...))
```

## Summarize

```{r}
# Calculate mean, median, minimum, maximum, and number of non-missing values.

one_variable |>
  summarize(
    n = n(),
    mean_value = mean(...),
    median_value = median(...),
    min_value = min(...),
    max_value = max(...)
  )
```

## Plot

```{r}
# Make a histogram.

ggplot(
  one_variable,
  aes(x = ...)
) +
  geom_histogram(binwidth = ...) +
  labs(
    title = "...",
    x = "...",
    y = "Count"
  ) +
  theme_minimal()
```

## Change The Bin Width

```{r}
# Make the same histogram with a different bin width.

ggplot(
  one_variable,
  aes(x = ...)
) +
  geom_histogram(binwidth = ...) +
  labs(
    title = "...",
    x = "...",
    y = "Count"
  ) +
  theme_minimal()
```

## Boxplot By Group

Choose one numeric variable and one categorical grouping variable.

Suggestion: use `age_group` for the grouping variable.

```{r}
# Make a boxplot to compare spread across groups.

boxplot_data <- dat |>
  select(age_group, ...) |>
  filter(!is.na(age_group)) |>
  filter(!is.na(...))

ggplot(
  boxplot_data,
  aes(x = age_group, y = ...)
) +
  geom_boxplot() +
  labs(
    title = "...",
    x = "Age group",
    y = "..."
  ) +
  theme_minimal()
```

The boxplot shows the median, the middle 50% of values, the whiskers, and points flagged by the default boxplot rule.

# Interpretation

Write 2-3 sentences:

- What does the histogram show about the shape of the variable?
- How did changing the bin width change what you could see?
- What does the boxplot show about spread across groups?
