---
title: "Boxplots, Violin Plots, And Outliers"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
if (!requireNamespace("DT", quietly = TRUE)) {
  install.packages("DT")
}

library(dplyr)
library(ggplot2)
```

# Goal

This notebook shows how to use summaries, boxplots, violin plots, and tables to understand spread and unusual values. We'll continue with the NHANES teaching file.

# Analytic Questions

- How much spread is there?
- How do boxplots connect to IQR fences?
- What do violin plots show that boxplots hide?
- Which values might deserve follow-up by a simple IQR rule?

# Load Data

```{r}
nhanes <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")

nhanes |>
  glimpse()
```

# Prepare Dietary Data

```{r}
diet <- nhanes |>
  filter(!is.na(mean_daily_energy_kcal))

diet |>
  select(respondent_id, age_years, gender, age_group, mean_daily_energy_kcal) |>
  head()
```

The dietary summaries are simple averages over available recall days. They are for exploratory teaching examples, not formal usual-intake estimation.

# Summarize Spread

Summaries can describe spread in a few numbers.

```{r}
energy_summary <- diet |>
  summarize(
    n = n(),
    mean_energy = mean(mean_daily_energy_kcal),
    median_energy = median(mean_daily_energy_kcal),
    min_energy = min(mean_daily_energy_kcal),
    max_energy = max(mean_daily_energy_kcal),
    iqr_energy = IQR(mean_daily_energy_kcal),
    sd_energy = sd(mean_daily_energy_kcal)
  )

energy_summary
```

The range shows the full span. The IQR describes the middle 50% of values. The standard deviation describes spread around the mean.

These summaries are useful, but they do not show the shape of the tails. That is why we plot too.

# Screen For Unusual Values

Here, we'll calculate quartiles and IQR fences ourself. `ggplot2` will do this for us, but it's always useful to remember how these work.

```{r}
energy_cutoffs <- diet |>
  summarize(
    q1 = quantile(mean_daily_energy_kcal, 0.25),
    q3 = quantile(mean_daily_energy_kcal, 0.75),
    iqr = IQR(mean_daily_energy_kcal),
    lower_cutoff = q1 - 1.5 * iqr,
    upper_cutoff = q3 + 1.5 * iqr
  )

energy_cutoffs
```

We calculate two fences/cutoffs. For this question, we'll focus on the high end of values, so we flag only values above the upper cutoff.

```{r}
diet_flagged <- diet |>
  mutate(
    high_energy_outlier = mean_daily_energy_kcal > energy_cutoffs$upper_cutoff
  )

diet_flagged |>
  count(high_energy_outlier)
```

```{r}
diet_not_flagged <- diet_flagged |>
  filter(!high_energy_outlier)

diet_high_flagged <- diet_flagged |>
  filter(high_energy_outlier)
```

These are the same basic fences that `ggplot2` uses by default when it draws a boxplot. The default boxplot rule is:

- lower fence: Q1 - 1.5 x IQR
- upper fence: Q3 + 1.5 x IQR

Points beyond those fences are drawn as outlier points.

# Inspect Flagged Rows

Flagged values are not automatically errors. Often they're perfectly natural results of the natural/social world.

```{r}
diet_flagged |>
  filter(high_energy_outlier) |>
  select(respondent_id, age_years, gender, age_group, dietary_days, mean_daily_energy_kcal) |>
  arrange(desc(mean_daily_energy_kcal)) |>
  head(10)
```

## Compare With A Box Plot

`ggplot2` makes it very easy to create boxplots.

The box shows the middle 50% of the data. The line inside the box is the median. The points beyond the whiskers use the same 1.5 IQR idea we calculated above.

Start with a one-variable boxplot. There is no grouping variable here, so every participant is part of the same distribution.

```{r}
ggplot(
  diet,
  aes(y = mean_daily_energy_kcal)
) +
  geom_boxplot() +
  labs(
    title = "Mean Daily Energy Intake",
    x = NULL,
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

Now draw the 1.5 IQR upper fence that we calculated earlier. Everything above the red line is beyond the 1.5 IQR upper fence.

```{r}
ggplot(
  diet,
  aes(y = mean_daily_energy_kcal)
) +
  geom_boxplot() +
  geom_hline(
    yintercept = energy_cutoffs$upper_cutoff,
    color = "firebrick",
    linewidth = 1
  ) +
  labs(
    title = "Mean Daily Energy Intake With 1.5 IQR Upper Fence",
    x = NULL,
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

The line makes the math visible: `geom_boxplot()` is using the same kind of fence we calculated by hand.

We can also make grouped boxplots.

```{r}
p1 <- ggplot(
  diet,
  aes(x = age_group, y = mean_daily_energy_kcal)
) +
  geom_boxplot()

p1
```

```{r}
p2 <- p1 +
  labs(
    title = "Mean Daily Energy Intake By Age Group",
    x = "Age group",
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()

p2
```

## Change The Boxplot Rule

The default `coef` is `1.5`, which means the whiskers extend to values within 1.5 IQRs of the box.

We can change that. A larger `coef` makes fewer points appear as outliers because the whiskers can extend farther.

First, calculate the 3.0 IQR fences from the same quartiles and IQR.

```{r}
energy_cutoffs_3 <- diet |>
  summarize(
    q1 = quantile(mean_daily_energy_kcal, 0.25),
    q3 = quantile(mean_daily_energy_kcal, 0.75),
    iqr = IQR(mean_daily_energy_kcal),
    lower_cutoff = q1 - 3 * iqr,
    upper_cutoff = q3 + 3 * iqr
  )

energy_cutoffs_3
```

Now set `coef = 3` in the one-variable boxplot and draw the 3.0 IQR upper fence.

```{r}
ggplot(
  diet,
  aes(y = mean_daily_energy_kcal)
) +
  geom_boxplot(coef = 3) +
  geom_hline(
    yintercept = energy_cutoffs_3$upper_cutoff,
    color = "firebrick",
    linewidth = 1
  ) +
  labs(
    title = "Mean Daily Energy Intake With 3.0 IQR Upper Fence",
    x = NULL,
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

```{r}
ggplot(
  diet,
  aes(x = age_group, y = mean_daily_energy_kcal)
) +
  geom_boxplot(coef = 3) +
  labs(
    title = "Mean Daily Energy Intake By Age Group",
    x = "Age group",
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

## Change The Outlier Points

We can also keep the default rule but change how the outlier points look.

```{r}
ggplot(
  diet,
  aes(x = age_group, y = mean_daily_energy_kcal)
) +
  geom_boxplot(
    outlier.color = "firebrick",
    outlier.shape = 16,
    outlier.size = 2
  ) +
  labs(
    title = "Mean Daily Energy Intake By Age Group",
    x = "Age group",
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

## Plot The Values

We can hide the outlier points drawn by the boxplot and then add all the individual values, using jitter to spread them out.

This helps us see how the full distribution looks instead of only seeing the points flagged by the boxplot rule.

In `ggplot2`, layer order matters. Layers added later are drawn on top of earlier layers.

First, draw the boxplot and put the jittered points on top.

```{r}
ggplot(
  diet_flagged,
  aes(x = age_group, y = mean_daily_energy_kcal)
) +
  geom_boxplot(outlier.shape = NA, fill = "white", alpha = 0.8) +
  geom_jitter(
    data = diet_not_flagged,
    color = "gray60",
    width = 0.15,
    alpha = 0.35,
    size = 0.8
  ) +
  geom_jitter(
    data = diet_high_flagged,
    aes(color = "Flagged"),
    width = 0.15,
    alpha = 0.8,
    size = 0.8
  ) +
  scale_color_manual(
    values = c("Flagged" = "firebrick"),
    name = NULL
  ) +
  labs(
    title = "Mean Daily Energy Intake By Age Group",
    x = "Age group",
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

Now reverse the order: draw the jittered points first and put the boxplot on top.

```{r}
ggplot(
  diet_flagged,
  aes(x = age_group, y = mean_daily_energy_kcal)
) +
  geom_jitter(
    data = diet_not_flagged,
    color = "gray60",
    width = 0.15,
    alpha = 0.35,
    size = 0.8
  ) +
  geom_jitter(
    data = diet_high_flagged,
    aes(color = "Flagged"),
    width = 0.15,
    alpha = 0.8,
    size = 0.8
  ) +
  geom_boxplot(outlier.shape = NA, fill = "white", alpha = 0.8) +
  scale_color_manual(
    values = c("Flagged" = "firebrick"),
    name = NULL
  ) +
  labs(
    title = "Mean Daily Energy Intake By Age Group",
    x = "Age group",
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

The data are the same in both plots. The layer order changes which part of the display visually sits on top.

Adding points can help add useful context about the plot, such as how many points are in each category.

# Show The Flagged Rows In A Table

When appropriate, we can show collaborators the rows behind the flagged values.

Here we use the wider 3.0 IQR rule and arrange the table by the energy value.

```{r}
energy_3_iqr_flagged <- diet |>
  filter(mean_daily_energy_kcal > energy_cutoffs_3$upper_cutoff) |>
  arrange(desc(mean_daily_energy_kcal)) |>
  select(
    respondent_id,
    age_years,
    gender,
    age_group,
    dietary_days,
    mean_daily_energy_kcal
  )

DT::datatable(
  energy_3_iqr_flagged,
  rownames = FALSE,
  options = list(pageLength = 10)
)
```

Interactive tables can be useful for review, but use them carefully. Do not share sensitive row-level data unless it is appropriate for the setting and audience.

# Violin Plots

Violin plots use smoothing to show the shape of a distribution within each group. They let you see the distribution of a measure through the density, and get a better sense of the range/outliers through the boxplot.

```{r}
ggplot(
  diet,
  aes(x = age_group, y = mean_daily_energy_kcal)
) +
  geom_violin(fill = "gray85", color = "gray40") +
  labs(
    title = "Mean Daily Energy Intake By Age Group",
    x = "Age group",
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

The violin plot makes the width of each distribution visible, but it hides the median and the middle 50% of values unless we add another layer.

```{r}
ggplot(
  diet,
  aes(x = age_group, y = mean_daily_energy_kcal)
) +
  geom_violin(fill = "gray85", color = "gray40") +
  geom_boxplot(
    width = 0.12,
    outlier.shape = NA,
    fill = "white",
    color = "black",
    linewidth = 0.7
  ) +
  labs(
    title = "Mean Daily Energy Intake By Age Group",
    x = "Age group",
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

The boxplot layer adds the median and middle spread back onto the smoothed shape.

```{r}
ggplot(
  diet_flagged,
  aes(x = age_group, y = mean_daily_energy_kcal)
) +
  geom_violin(fill = "gray90", color = "gray50") +
  geom_jitter(
    data = diet_not_flagged,
    color = "gray60",
    width = 0.15,
    alpha = 0.25,
    size = 0.7
  ) +
  geom_jitter(
    data = diet_high_flagged,
    aes(color = "Flagged"),
    width = 0.15,
    alpha = 0.8,
    size = 0.8
  ) +
  scale_color_manual(
    values = c("Flagged" = "firebrick"),
    name = NULL
  ) +
  labs(
    title = "Mean Daily Energy Intake By Age Group",
    x = "Age group",
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()
```

This is not automatically better than a boxplot. Above is another view of a density plot, with plotted values, and highlighted outliers.

# Takeaways

- Unusual values can be errors, true rare values, or signs that the variable needs context.
- The default boxplot rule of 1.5 IQR is a screening tool, not a final judgment. Use your scientific judgment to inform what you would consider an outlier.
- Visualization helps show whether unusual values are isolated or part of a tail.
- Try out different methods to see what is the most useful for you.
