---
title: "2. Setup Check"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Use this notebook to confirm that R, `ggplot2`, and Quarto rendering work on your computer.

Run each code chunk. Then render the notebook to HTML.

# Check 1: R Works

Run this chunk.

```{r}
2 + 2
```

You should see `4` in the output.

# Check 2: Data Works

R includes a small built-in dataset called `mtcars`. We can use it to test data work without downloading anything else.

```{r}
mtcars |>
  head()
```

# Check 3: Summaries Work

Now summarize the data.

```{r}
mtcars |>
  summarise(
    rows = n(),
    average_mpg = mean(mpg),
    minimum_mpg = min(mpg),
    maximum_mpg = max(mpg)
  )
```

Question: how many rows are in `mtcars`?

> Write your answer here.

# Check 4: ggplot2 Works

Now make a simple plot with `ggplot2`.

```{r}
p1 <- ggplot(mtcars, aes(x = mpg))

p1
```

The blank plot is expected. We have provided data and an x-axis mapping, but no geometric object yet.

Add a histogram.

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

Add labels.

```{r}
p3 <- p2 +
  labs(
    title = "Distribution Of Fuel Efficiency",
    subtitle = "Built-in mtcars dataset",
    x = "Miles per gallon",
    y = "Number of cars"
  )

p3
```

Add a theme.

```{r}
p4 <- p3 +
  theme_minimal()

p4
```

Question: what does this plot show?

> Write one or two sentences here.

# Check 5: Render Works

Render this notebook to HTML.

After rendering, answer these questions:

- Did the notebook render successfully?
- Where did the HTML file appear?
- Did the plot appear in the rendered HTML file?

Notes:

> Write any setup notes or error messages here.
